---
title: "Statistical Report"
author: "Katrina Li"
date: "2023-02-12"
output: 
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: cosmo
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(forcats)
library(ggplot2)
library(lme4)
library(lmerTest)
library(stringr)
library(sjPlot)
library(emmeans)
library(forcats)
```

# Cantonese (HLVC dataset)

We read two data files: 

```{r read data, echo=TRUE}
can_original <- read.delim("processed_canCHN.tsv", header = F, sep ="\t")
can_freq_original <- read.delim("processed_canCHN_freq.tsv",header = F, sep ="\t")
```

We then clean the datasets and merge the two, generating `can.regdata` used in the regression model.
```{r clean dataset}
can <- can_original %>%
  rename(
    sent_word = V1,
    sent_cn = V2,
    langs = V3,
    pos =V4,
    tone_initial = V5,
    tone_last = V6,
    en_cn = V7,
    info_ec_en = V8,
    cn_en = V9,
    info_ce_en = V10,
    cn_null_en = V11,
    info_cne_null = V12,
    info_cen_en = V13,
    en_null_cn = V14,
    info_enc_null = V15,
    info_enc_en = V16,
    filename = V17,
    sentence = V18,
    token =V19
  ) 

can_freq <- can_freq_original %>%
  rename(
    word = V1,
    freq = V2,
    lang = V3
  ) %>%
  mutate(
    lang = as.factor(lang),
    zipfreq = log10(freq*1000000)
  )


can_freq_formerge <- can_freq %>%
  filter(lang != "cn_dict" & lang != "tone") %>%
  mutate(
    langs = case_when(
      lang == "cn_char_dict" ~ "cn",
      lang == "en_dict" ~ "en",
      lang == "filler_dict" ~ "null"
    )
  )
  
can.combine <- can %>%
  left_join(can_freq_formerge, by = c("sent_cn" = "word", "langs" = "langs")) %>%
  mutate(
    cn_en = as.factor(cn_en),
    en_cn = as.factor(en_cn),
    freq = as.numeric(freq),
    zipfreq = as.numeric(zipfreq),
    POS = as.factor(pos),
    speaker = sub("\\_.*","",filename),
    SPEAKER = as.factor(speaker),
    generation = substr(speaker,2,2),
    GENERATION = if_else(generation == "2"|generation == "3", "23", generation),
    GENERATION = as.factor(GENERATION),
    gender = substr(speaker,3,3),
    GENDER = as.factor(gender),
    TOKEN = if_else(zipfreq > 4, sent_cn, "other"), # control the frequent words
    TOKEN = as.factor(TOKEN),
    filename = as.factor(filename)
  )

# check if any rows in the dataset are not merged
# can %>%
#   anti_join(can_freq_formerge, by = c("sent_cn" = "word", "langs" = "langs"))

# Regression data
can.regdata <- can.combine %>%
  filter(
    langs == "cn"
  ) %>%
  mutate(
    en_cn = as.factor(en_cn),
    cn_en = as.factor(cn_en),
    freq = as.numeric(freq),
    zipfreq = as.numeric(zipfreq),
    tone_initial = as.factor(tone_initial),
    tone_last = as.factor(tone_last),
    pos = if_else(pos == "V", "VERB", pos),
    POS = as.factor(pos),
    SPEAKER = as.factor(speaker),
    GENERATION = as.factor(GENERATION),
    GENDER = as.factor(gender),
    TOKEN = if_else(zipfreq > 4, sent_word, "other"), # control the frequent words
    TOKEN = as.factor(TOKEN)
  )

# Assign the reference level of factor variables
can.regdata$tone_last = as.factor(can.regdata$tone_last)
can.regdata$tone_last = relevel(can.regdata$tone_last, ref = "4")
can.regdata$tone_initial = as.factor(can.regdata$tone_initial)
can.regdata$tone_initial = relevel(can.regdata$tone_initial, ref = "4")
can.regdata$GENERATION = relevel(can.regdata$GENERATION, ref = "X")
can.regdata$GENDER = relevel(can.regdata$GENDER, ref = "M")
can.regdata$POS = relevel(can.regdata$POS, ref = "ADJ")
can.regdata$SPEAKER = droplevels(can.regdata$SPEAKER)
can.regdata$TOKEN = droplevels(can.regdata$TOKEN)
```

## Summary statistics of HLVC variables

### SPEAKER

balenced in gender and generation
```{r about speaker}
can.speaker <- distinct(can.regdata,SPEAKER,.keep_all = TRUE)
table(can.speaker$GENDER)
table(can.speaker$GENERATION)
table(can.speaker$GENERATION, can.speaker$GENDER)
```

### TOKEN

14 tokens with high frequency + 'other'
```{r}
nlevels(can.regdata$TOKEN)
table(can.regdata$TOKEN)
```

### Zipfreq

14 Chinese words has a zipf larger than 4, cover 30% of the tokens.

```{r}
mean(can.regdata$zipfreq)
sd(can.regdata$zipfreq)
nrow(can.regdata[can.regdata$zipfreq > 4,])/nrow(can.regdata)
```

### POS

```{r}
nlevels(can.regdata$POS)
table(can.regdata$POS)
```

## Regression models of Cantonese
### Cantonese-English code switching

**can.ce <- glmer(cn_en ~ tone_last + zipfreq + POS + GENDER + GENERATION + 
                                   (tone_last|SPEAKER) +
                                   (1|TOKEN),
                           data = can.regdata,
                           family = binomial("logit"),
                           control=glmerControl(optimizer="bobyqa", 
                           optCtrl=list(maxfun=2e5),
                           calc.derivs=FALSE), 
                           na.action=na.omit
                          )**
```{r include=FALSE}
can.ce <- readRDS("230205can_ce.glmer.final.rds")
```

```{r full model}
summary(can.ce)
```

**Post-hoc comparisons**
```{r post hoc}
emmeans(can.ce, list(pairwise~tone_last), type='response')
```

### English-Cantonese code-switching:

**can.ec <- glmer(en_cn ~ tone_initial + zipfreq + POS + GENDER + GENERATION + 
                                   (tone_initial|SPEAKER) +
                                   (1|TOKEN),
                           data = can.regdata,
                           family = binomial("logit"),
                           control=glmerControl(optimizer="bobyqa", 
                           optCtrl=list(maxfun=2e5),
                           calc.derivs=FALSE), 
                           na.action=na.omit
                          )**
                          
```{r include=FALSE}
can.ec <- readRDS("230205can_ec.glmer.final.rds")
```

Full model:
```{r}
summary(can.ec)
```

**Post-hoc comparison**
```{r}
emmeans(can.ec, list(pairwise~tone_initial), type='response')
```

# Mandarin (SEAME)
We read two data files: 

```{r echo=TRUE}
man_original <- read.delim("processed_man.tsv", header = F, sep ="\t")
man_freq_original <- read.delim("processed_man_freq.tsv",header = F, sep ="\t")
```

```{r read data and merge with frequency}
man <- man_original %>%
  rename(
    sent_word = V1,
    sent_cn = V2,
    langs = V3,
    pos =V4,
    tone_initial = V5,
    tone_last = V6,
    en_cn = V7,
    info_ec_en = V8,
    cn_en = V9,
    info_ce_en = V10,
    cn_null_en = V11,
    info_cne_null = V12,
    info_cen_en = V13,
    en_null_cn = V14,
    info_enc_null = V15,
    info_enc_en = V16,
    filename = V17,
    sentence = V18,
    token =V19,
    txtname = V20
  ) 

man_freq <- man_freq_original %>%
  rename(
    word = V1,
    freq = V2,
    lang = V3
  ) %>%
  mutate(
    lang = as.factor(lang),
    freq = as.numeric(freq),
    zipfreq = log10(freq*1000000)
  )
#summary(man_freq)

man_freq_formerge <- man_freq %>%
  filter(lang != "cn_dict" & lang != "tone") %>%
  mutate(
    langs = case_when(
      lang == "cn_char_dict" ~ "cn",
      lang == "en_dict" ~ "en",
      lang == "filler_dict" ~ "null"
    )
  )

  
man.combine <- man %>%
  left_join(man_freq_formerge, by = c("sent_cn" = "word", "langs" = "langs")) %>%
  mutate(
    #tone_last = as.factor(tone_last),
    #tone_initial = as.factor(tone_initial),
    cn_en = as.factor(cn_en),
    en_cn = as.factor(en_cn),
    freq = as.numeric(freq),
    zipfreq = as.numeric(zipfreq),
    POS = as.factor(pos),
    speaker = sub("\\_.*","",filename),
    SPEAKER = substring(speaker,1,5),
    GENDER = substring(SPEAKER,nchar(SPEAKER),nchar(SPEAKER)),
    TOKEN = if_else(zipfreq > 4, sent_cn, "other"), # control the frequent words
    TOKEN = as.factor(TOKEN),
    langgroup = gsub("^.*/","",txtname),
    LANGUAGE = substring(langgroup,5,7),
    LANGUAGE = as.factor(LANGUAGE),
    SPEAKER = as.factor(SPEAKER),
    GENDER = as.factor(GENDER)
  )


# Check the merge direction from the other
# man %>%
#   anti_join(man_freq_formerge, by = c("sent_cn" = "word"))

# Regression data
man.regdata <- man.combine %>%
  filter(
    langs == "cn"
  ) %>%
  mutate(
    en_cn = as.factor(en_cn),
    cn_en = as.factor(cn_en),
    freq = as.numeric(freq),
    zipfreq = as.numeric(zipfreq),
    tone_initial = as.factor(tone_initial),
    tone_last = as.factor(tone_last),
    POS = as.factor(pos),
    SPEAKER = as.factor(SPEAKER),
    LANGUAGE = as.factor(LANGUAGE),
    GENDER = as.factor(GENDER),
    TOKEN = if_else(zipfreq > 4, sent_word, "other"), # control the frequent words
    TOKEN = as.factor(TOKEN)
  )

# Assign the reference level of factor variables
man.regdata$tone_last = as.factor(man.regdata$tone_last)
man.regdata$tone_last = relevel(man.regdata$tone_last, ref = "2")
man.regdata$tone_initial = as.factor(man.regdata$tone_initial)
man.regdata$tone_initial = relevel(man.regdata$tone_initial, ref = "2")
man.regdata$GENDER = relevel(man.regdata$GENDER, ref = "m")
man.regdata$LANGUAGE = relevel(man.regdata$LANGUAGE, ref = "man")
man.regdata$POS = relevel(man.regdata$POS, ref = "ADJ")
man.regdata$SPEAKER = droplevels(man.regdata$SPEAKER)
man.regdata$TOKEN = droplevels(man.regdata$TOKEN)
```

## Summary statistics of SEAME variables

### SPEAKER

balanced for language dominance and gender
```{r speaker}
man.speaker <- distinct(man.combine,SPEAKER,.keep_all = TRUE)
table(man.speaker$LANGUAGE, man.speaker$GENDER)
table(man.speaker$LANGUAGE, man.speaker$GENDER)
```

### TOKEN

24 tokens with high frequency  + 'other'

```{r}
nlevels(man.regdata$TOKEN)
table(man.regdata$TOKEN)
```

### Zipfreq

23 Chinese words has a zipf larger than 4, cover 46% of the tokens.

```{r}
mean(man.regdata$zipfreq)
sd(man.regdata$zipfreq)
nrow(man.regdata[man.regdata$zipfreq > 4,])/nrow(man.regdata)
```

### POS
```{r}
nlevels(man.regdata$POS)
table(man.regdata$POS)
```

## Regression models of Mandarin
### Mandarin-English code switching
**man.me <- glmer(cn_en ~ tone_last + zipfreq + POS + GENDER + LANGUAGE + 
                                   (tone_last|SPEAKER) +
                                   (1|TOKEN),
                           data = man.regdata,
                           family = binomial("logit"),
                           control=glmerControl(optimizer="bobyqa", 
                           optCtrl=list(maxfun=2e5),
                           calc.derivs=FALSE), 
                           na.action=na.omit
                          )**
                          
```{r include=FALSE}
man.me <- readRDS("230205man_me.glmer.final.rds")
man.em <- readRDS("230205man_em.glmer.final.rds")
```

```{r}
summary(man.me)
```

**Post-hoc comparisons**
```{r}
emmeans(man.me, list(pairwise~tone_last), type='response')
```

### English-Mandarin code switching
**man.em <- glmer(en_cn ~ tone_initial + zipfreq + POS + GENDER + LANGUAGE + 
                                   (tone_initial|SPEAKER) +
                                   (1|TOKEN),
                           data = man.regdata,
                           family = binomial("logit"),
                           control=glmerControl(optimizer="bobyqa", 
                           optCtrl=list(maxfun=2e5),
                           calc.derivs=FALSE), 
                           na.action=na.omit
                          )**

```{r}
summary(man.em)
```

**Post-hoc comparisons**
```{r}
emmeans(man.em, list(pairwise~tone_initial), type='response')
```

# Vietnamese (CanVEC)
We read three data files: 

speakerprofile is extracted from Nguyen 's thesis (2020).

```{r read the files, echo=TRUE}
vie_original <- read.delim("processed_canvec.tsv", header = F, sep ="\t")
vie_freq_original <- read.delim("processed_vie_freq.tsv",header = F, sep ="\t")
vie_speaker_original <- read.delim("speakerprofile.csv", header = F, sep = ",") 
```

```{r read viet data and merge with frequency}
vie <- vie_original %>%
  rename(
    sent_word = V1,
    langs = V2,
    pos =V3,
    tone = V4,
    en_vn = V5,
    info_ev_en = V6,
    vn_en = V7,
    info_ve_en = V8,
    vn_null_en = V9,
    info_vne_null = V10,
    info_vne_en = V11,
    en_null_vn = V12,
    info_env_null = V13,
    info_env_en = V14,
    speaker = V15,
    generation = V16,
    sentence = V17,
    token =V18
  ) %>%
  mutate(speaker2 = case_when(
  speaker == "Dany2" ~ "Dany",
  speaker == "Jess2" ~ "Jess",
  speaker == "Henry2" ~ "Henry",
  speaker == "Tom2" ~ "Tom",
  speaker == "Tim2" ~ "Tim",
  speaker == "Tressie2" ~ "Tressie",
  speaker == "﻿Billy" ~ "Billy",
  TRUE ~ speaker
  ),
  #langs = if_else(sent_word == "nam", "@non", langs),
  pos = if_else(pos=="Z","X",pos),
  tone.list = gsub("\\[|\\]", "", tone),
  tone.list = gsub(" ","", tone.list),
  tone_initial = substr(tone.list, 1,1),
  tone_last = substr(tone.list,nchar(tone.list),nchar(tone.list)),
  leng = nchar(tone.list))

vie_freq <- vie_freq_original %>%
  rename(
    word = V1,
    freq = V2,
    lang = V3
  ) %>%
  mutate(
    lang = as.factor(lang),
    freq = as.numeric(freq),
    zipfreq = log10(freq*1000000)
  )

vie_freq_formerge <- vie_freq %>%
  filter(lang != "tone") %>%
  mutate(
    langs = case_when(
      lang == "en" ~ "@eng",
      lang == "vn" ~ "@vie",
      lang == "filler_dict" ~ "@non"
    )
  ) %>%
  select(-lang)


vie_speaker <- vie_speaker_original %>%
  rename(
    RecordingNo = V1,
    Pseudonym = V2,
    Generation = V3,
    Gender = V4,
    Year.of.Birth = V5,
    Occupation = V6
  ) 

vie_speaker_formerge <- vie_speaker %>%
  select(Pseudonym, Gender, Year.of.Birth)


vie.combine <- vie %>%
  inner_join(vie_freq_formerge, by = c("sent_word" = "word", "langs" = "langs")) %>%
  inner_join(vie_speaker_formerge, by = c("speaker2" = "Pseudonym")) %>%
  mutate(
    vn_en = as.factor(vn_en),
    en_vn = as.factor(en_vn),
    freq = as.numeric(freq),
    zipfreq = as.numeric(zipfreq),
    POS = as.factor(pos),
    SPEAKER = as.factor(speaker2),
    GENERATION = as.factor(generation),
    GENDER = as.factor(Gender),
    TOKEN = if_else(zipfreq > 4, sent_word, "other"), # control the frequent words
    TOKEN = as.factor(TOKEN),
  )


# Check the merge direction from the other
# vie %>%
#   anti_join(vie_freq_formerge, by = c("sent_word" = "word"))
# vie %>%
#   anti_join(vie_speaker_formerge, by = c("speaker2" ="Pseudonym"))

# Regression data
vie.regdata <- vie.combine %>%
  filter(
    langs == "@vie",
    tone != "0"
  ) %>%
  mutate(
    en_vn = as.factor(en_vn),
    vn_en = as.factor(vn_en),
    freq = as.numeric(freq),
    zipfreq = as.numeric(zipfreq),
    tone_initial = as.factor(tone_initial),
    tone_last = as.factor(tone_last),
    POS = as.factor(pos),
    SPEAKER = as.factor(speaker2),
    GENERATION = as.factor(generation),
    GENDER = as.factor(Gender),
    TOKEN = if_else(zipfreq > 4, sent_word, "other"), # control the frequent words
    TOKEN = as.factor(TOKEN)
  )

vie.regdata$tone_initial = as.factor(vie.regdata$tone_initial)
vie.regdata$tone_initial = relevel(vie.regdata$tone_initial, ref = "4")
vie.regdata$tone_last = as.factor(vie.regdata$tone_last)
vie.regdata$tone_last = relevel(vie.regdata$tone_last, ref = "4")
vie.regdata$GENERATION = relevel(vie.regdata$GENERATION, ref = "1")
vie.regdata$GENDER = relevel(vie.regdata$GENDER, ref = "M")
vie.regdata$POS = relevel(vie.regdata$POS, ref = "ADJ")
vie.regdata$SPEAKER = droplevels(vie.regdata$SPEAKER)
vie.regdata$TOKEN = droplevels(vie.regdata$TOKEN)
```

### SPEAKER
2 speakers are excluded because they only speak English.
```{r speaker profile}
nlevels(vie.combine$SPEAKER)
nlevels(vie.regdata$SPEAKER) 

vie.speaker <- distinct(vie.regdata,SPEAKER,.keep_all = TRUE)
table(vie.speaker$GENERATION, vie.speaker$GENDER)
```

### TOKEN
14 tokens with high frequency + 'other'
```{r token}
nlevels(vie.regdata$TOKEN)
table(vie.regdata$TOKEN)
```

### Zipfreq
14 Chinese words has a zipf larger than 4, cover 27% of the tokens.

```{r}
mean(vie.regdata$zipfreq)
sd(vie.regdata$zipfreq)
nrow(vie.regdata[vie.regdata$zipfreq > 4,])/nrow(vie.regdata)
```


### POS
```{r POS}
nlevels(vie.regdata$POS)
table(vie.regdata$POS)
```

## Regression models of Vietnamese

```{r include=FALSE}
vie.ve <- readRDS("230205vie_ve.glmer.final.rds")
vie.ev <- readRDS("230205vie_ev.glmer.final.rds")
```

### Vietnamese-English code switching

**vie.ve <- glmer(vn_en ~ tone_last + zipfreq + POS + GENDER + GENERATION + 
                                   (tone_last|SPEAKER) +
                                   (1|TOKEN),
                           data = vie.regdata,
                           family = binomial("logit"),
                           control=glmerControl(optimizer="bobyqa", 
                           optCtrl=list(maxfun=2e5),
                           calc.derivs=FALSE), 
                           na.action=na.omit
                          )**
```{r}
summary(vie.ve)
```

**Post-hoc comparisons**

```{r}
emmeans(vie.ve, list(pairwise~tone_last), type='response')
```
### English-Vietnamese code switching
**vie.ev <- glmer(en_vn ~ tone_initial + zipfreq + POS + GENDER + GENERATION + 
                                   (tone_initial|SPEAKER) +
                                   (1|TOKEN),
                           data = vie.regdata,
                           family = binomial("logit"),
                           control=glmerControl(optimizer="bobyqa", 
                           optCtrl=list(maxfun=2e5),
                           calc.derivs=FALSE), 
                           na.action=na.omit
                          )**
```{r}
summary(vie.ev)
```
```{r}
emmeans(vie.ev, list(pairwise~tone_initial), type='response')
```

